---
title: "Lab 6"
author: "Kelbi Hall"
date: "2025-03-01"
format: html
execute:
  echo: true
---

```{r}
library(tidyverse)
library(readxl)
library(ggthemes)
library(tidymodels)
library(vip)
library(baguette)
library(glue)
```

## Question#1

### What does `zero_q_freq` represent?

The CAMELS dataset doc says that `zero_q_freq` is the fraction of days with zero stream flow over the observational period. So this means it is a measure of how often a stream at a given site runs dry.

```{r}
clim <- read_delim("data/camels_clim.txt", delim = ";")
geol <- read_delim("data/camels_geol.txt", delim = ";")
hydro <- read_delim("data/camels_hydro.txt", delim = ";")
soil <- read_delim("data/camels_soil.txt", delim = ";")
topo <- read_delim("data/camels_topo.txt", delim = ";")
vege <- read_delim("data/camels_vege.txt", delim = ";")

camels <- reduce(list(clim, geol, hydro, soil, topo, vege), full_join, by = "gauge_id")

ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = q_mean)) +
  scale_color_gradient(low = "pink", high = "dodgerblue") +
  ggthemes::theme_map() +
  labs(title = "CAMELS Sites Colored by Mean Flow (q_mean)",
       x = "Longitude", y = "Latitude",
       color = "Mean Flow")

```

## Question #2

```{r}
library(patchwork)

map_p <- ggplot(camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray60") +
  geom_point(aes(color = p_mean), size = 2) +
  scale_color_viridis_c(option = "D") +
  coord_fixed(1.3) +  # <- this is the fix!
  ggthemes::theme_map() +
  labs(title = "Mean Precipitation (p_mean)",
       color = "mm/day")

map_a <- ggplot(camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray60") +
  geom_point(aes(color = aridity), size = 2) +
  scale_color_viridis_c(option = "C") +
  coord_fixed(1.3) +  # <- fix here too
  ggthemes::theme_map() +
  labs(title = "Aridity Index (PET / P)",
       color = "Aridity")

map_p + map_a

```

## Model Building

```{r}
library(tidymodels)


camels_mod <- camels %>%
  select(where(is.numeric)) %>%
  drop_na()


set.seed(330)
split <- initial_split(camels_mod, prop = 0.8)
train <- training(split)
test <- testing(split)


rec <- recipe(q_mean ~ ., data = train) %>%
  step_normalize(all_predictors())

```

```{r}
lm_mod <- linear_reg() %>% set_engine("lm")
rf_mod <- rand_forest(mode = "regression") %>% set_engine("ranger")
xgb_mod <- boost_tree(mode = "regression") %>% set_engine("xgboost")

```

```{r}
models <- workflow_set(
  preproc = list(base = rec),
  models = list(LM = lm_mod, RF = rf_mod, XGB = xgb_mod)
)

resamples <- vfold_cv(train, v = 5)


results <- models %>%
  workflow_map("fit_resamples", resamples = resamples, metrics = metric_set(rmse, rsq))

```

```{r}
collect_metrics(results) %>% 
  select(wflow_id, .metric, mean, std_err)
```

```{r}
final_lm <- workflow() %>%
  add_recipe(rec) %>%
  add_model(lm_mod) %>%
  fit(data = train)


lm_preds <- predict(final_lm, test) %>%
  bind_cols(test)

metrics(lm_preds, truth = q_mean, estimate = .pred)


ggplot(lm_preds, aes(x = q_mean, y = .pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Test Set: Predicted vs Observed q_mean (Linear Model)",
       x = "Observed q_mean",
       y = "Predicted q_mean")

```

## Question #3

```{r}
library(baguette)

nn_mod <- bag_mlp(mode = "regression") %>%
  set_engine("nnet")

```

```{r}
models_all <- workflow_set(
  preproc = list(base = rec),
  models = list(
    LM = lm_mod,
    RF = rf_mod,
    XGB = xgb_mod,
    NN = nn_mod
  )
)

# Fit with cross-validation
results_all <- models_all %>%
  workflow_map("fit_resamples", resamples = resamples, metrics = metric_set(rmse, rsq))

```

```{r}
collect_metrics(results_all) %>%
  select(wflow_id, .metric, mean, std_err)
```

After adding a neural net model, I compared linear, RF, XGBoost, and NN. Linear regression still performed the best overall with the lowest RMSE and highest R². The neural net and XGBoost were close behind, but the simplicity and strong performance of the linear model makes it the best choice for this dataset.
